The prime of their life - the halflife of acting talent

### 2017-09-03 01_halflife-of-acting-talent/article.Rmd
Datatables tables

There are significant confounding factors that might prevent us from answering this question difinitively (which I'll discuss later) but let's dig into the data.

@@

# Packages
library(readr)
library(dplyr)
library(tidyr)
library(gender)
library(stringr)
library(lubridate)
library(highcharter)
library(purrr)
library(htmltools)
library(pander)

@@

dict %>% pandoc.table(style = "rmarkdown")

@@

Let's have a look at the films data set
```{r results='asis'}
# Load in films data from Github Repo
films = read_csv("https://raw.githubusercontent.com/TMBish/lab_posts/master/01_halflife-of-acting-talent/data/film-data.csv") %>% filter(!is.na(title))

films %>% 
  filter(title %in% c("The Departed", "Titanic", "No Country for Old Men", "Snatch.", "Superbad")) %>%
  pandoc.table(emphasize.strong.cols = 3, missing="?", style = "rmarkdown", justify="left",split.cells=15)

@@

Unfortunately, because the Rotten Tomatoes development team are a bunch of politically correct SJWs, gender isn't provided on their film bio pages so we'll have to be creative. The `gender` R package allows you to infer a person's gender from their first name and DOB. The result is a probabilistic estimate of that individuals most likely gender given their first name and DOB.

@@

We should sense check the gender mapping to make sure the package is making logical inferences. Let's take a random 10 person sample:

```{r results='asis'}
genders %>%
  sample_n(10) %>%
  select(first_name, dob_year, gender) %>%
  pandoc.table(emphasize.strong.cols = 3, style = "rmarkdown", justify="left", split.cells=15)
```

@@

Well, as it turns out, it depends on which one of the above groups you're in. To remind you, our groups are: directors (predominantly male as seen above), male actors, and female actors. For all these individuals we calculate their age at the time they were involved in a realeased film and plot the distribution within each group.

@@

# Create chart
age_dist_chart = 
  highchart() %>%
  hc_add_theme(tmbish) %>%
  hc_chart(
    animation = list(duration = 2000)
  ) %>%
  hc_xAxis(
    categories = chart_data$age_at_production,
    title = list(text = "Age at Production")
  ) %>%
  hc_yAxis(
    title = list(text = "Probablity Density"),
    labels = list(formatter = JS("function(){return(Math.round(this.value * 1000) / 10 + '%')}"))
  ) %>%
  hc_title(text = "Age at Production") %>%
  hc_subtitle(text = "Comparing the distribution of film professionals for age at the time of film productions") %>%
  hc_add_series(
    name = "Director", 
    type = "spline",
    data = chart_data$director,
    marker = list(enabled = FALSE)
  ) %>%
  hc_add_series(
    name = "Male Actor",
    type = "spline",
    data = chart_data$male_actor,
    marker = list(enabled = FALSE)
  ) %>%
  hc_add_series(
    name = "Female Actor",
    type = "spline",
    data = chart_data$female_actor,
    zIndex = -10
    #marker = list(enabled = FALSE)
    #fillOpacity = 0.5
  )

@@

Our ultimate question is: do actors and directors improve or deteriorate as they get older. To investigate this hypothesis we must make a couple of assumptions. Since we don't have individual actor or director performance metrics (like we have in sports) we'll have to assume an actor or director can have a material affect on the quality of the moviesthey appear in. There are obviously many other factors that go into the quality of a film but this seems like a relatively safe assumption.

Further - for actors - we'll have to limit the dataset to actors that played an important role in the film. To implement an `actor performance == film quality` approach the actors need to be noteworthy enough to be able to affect the overall quality of the film. So while Michael Cera's cameo in **This is the End** was high quality it can't be said that he influenced the rotten tomatoes rating of the film on his own.

@@

There's a few other things we want to control for here. I don't necessarily want to plot raw film critical rating because this might be prone to idiosyncracies within groups or ages - for example, high performing directors might generally wait until they feel experienced enough to direct their first film.

Instead we'll plot, for each individual and film, how well they have perfomed relative to that individual's career average (the average tomatometer of all that individual's films). To illustrate this approach let's look at Denzel Washington's films. Denzel has a healthy mean tomatometer of 68%.

```{r echo=FALSE, results='asis'}

master %>% 
  filter(
    !(role %in% c("actor_4","actor_5", "actor_6")),
    name == "Denzel Washington",
    between(age_at_production, 15,85)
  ) %>%
  group_by(name) %>% 
  mutate(
    average_tomato  = round(mean(tomatometer))
  ) %>% 
  ungroup() %>%
  mutate(
    tomato_index = (tomatometer - average_tomato) / average_tomato
  ) %>%
  sample_n(10) %>%
  arrange(age_at_production) %>%
  select(title, year, age_at_production, tomatometer, average_tomato, tomato_index) %>%
  mutate(tomato_index = scales::percent(tomato_index)) %>%
  pandoc.table(emphasize.strong.cols = 6, style = "rmarkdown", justify="left")

So based on this 10 film sample Denzel had a solid early 40s where his films (of the likes **Hurricane** and **Courage Under Fire**) average a rotten tomatoes rating about 20% above his career average. He looks like he dropped off after 50 with a few flops like **Déjà Vu** (which I didn't mind) and **2 Guns**. Look I'm not calling him washed but he hasn't done anything good for a while.

@@

To help visualise this approach let's take 12 of the top 20 actors by volume and visualise average their film rating performance by age. De Niro is a great example of someone who starred in increasingly poor films as he got older - clearly visible in the chart below.. Obviously this probably has more to do with role selection than deteriorating skills (we'll discuss this later) but it's interesting to visualise nonetheless.

@@

## The holistic trend

@@

While blah blah blah

>>>>

There are some confounding factors that might prevent us from answering this question difinitively (which I'll discuss later) but let's dig into the data.

@@

# Packages
library(readr)
library(dplyr)
library(tidyr)
library(gender)
library(stringr)
library(lubridate)
library(highcharter)
library(purrr)
library(htmltools)
library(pander)
library(DT)

@@

dict %>% pandoc.table(justify='left')

@@

Let's have a look at the films data set. Click the blue `+` button to view the extra fields in the dataset (the director and top 6 actors in the film).

<br>
```{r results='asis'}
# Load in films data from Github Repo
films = read_csv("https://raw.githubusercontent.com/TMBish/lab_posts/master/01_halflife-of-acting-talent/data/film-data.csv") %>% filter(!is.na(title))

films %>% 
  filter(title %in% c("The Departed", "Titanic", "No Country for Old Men", "Snatch.", "Superbad")) %>%
  datatable(
    style = 'bootstrap',
    class = 'table-hover',
    rownames=FALSE,
    autoHideNavigation = TRUE,
    extensions = 'Responsive'

  )

```
<br>

@@

Unfortunately, because the Rotten Tomatoes development team are a bunch of politically correct SJWs, gender isn't provided on their film bio pages so we'll have to be creative. The `gender` R package allows you to infer a person's gender from their first name and DOB. The result is a probabilistic estimate of that individuals most likely gender given their first name and the year they were born.

@@

Pretty cool, but we should sense check the gender mapping to make sure the package is making logical guesses. Let's take a random 10 person sample:

```{r results='asis'}
genders %>%
  sample_n(10) %>%
  select(first_name, dob_year, gender) %>%
  datatable(
  style = 'bootstrap',
  class = 'table-hover',
  rownames=FALSE,
  autoHideNavigation = TRUE
  )
```

<bt>

@@

Well, as it turns out, it depends on which one of the above groups you're in. To remind you, our groups are: directors (predominantly male as seen above), male actors, and female actors. For all these individuals we calculate their age at the time they were credited in a film release and plot the distribution within each group.

@@

# Create chart
age_dist_chart = 
  highchart() %>%
  hc_add_theme(tmbish) %>%
  hc_chart(
    animation = list(duration = 2000)
  ) %>%
  hc_xAxis(
    categories = chart_data$age_at_production,
    title = list(text = "Age at Production")
  ) %>%
  hc_yAxis(
    title = list(text = "Probablity Density"),
    labels = list(formatter = JS("function(){return(Math.round(this.value * 1000) / 10 + '%')}"))
  ) %>%
  hc_title(text = "Age at Production") %>%
  hc_subtitle(text = "Comparing the distribution of film professionals for age at the time of film productions") %>%
  hc_add_series(
    name = "Director", 
    type = "spline",
    visible= FALSE,
    data = chart_data$director,
    marker = list(enabled = FALSE)
  ) %>%
  hc_add_series(
    name = "Male Actor",
    type = "spline",
    visible= FALSE,
    data = chart_data$male_actor,
    marker = list(enabled = FALSE)
  ) %>%
  hc_add_series(
    name = "Female Actor",
    type = "spline",
    data = chart_data$female_actor,
    zIndex = -10,
    marker = list(enabled = FALSE)
  ) %>%
  hc_tooltip(
    formatter = JS("function(){return('<b> Series: </b>' + point.key + '<br> <b> Age: </b>' + this.x + '<b> Value: </b>' + this.y )}")
  )

@@

Our ultimate question is: do actors and directors improve or deteriorate as they get older. To investigate this hypothesis we must make a couple of assumptions. Since we don't have individual actor or director performance metrics (like we have in sports) we'll have to assume an actor or director can have a material affect on the quality of the movies they appear in. There are obviously many other factors that go into the quality of a film but this seems like a relatively safe assumption.

Further - for actors - we'll have to limit the dataset to actors that played an important role in the film. To implement an `actor performance == film quality` approach the actors need to be influential enough to be able to affect the overall quality of the film. So while Michael Cera's cameo in **This is the End** was high quality, it can't be said that he influenced the rotten tomatoes rating of the film on his own.

@@

There's a few other things we want to control for here. I don't necessarily want to plot raw film critical rating because this might be prone to idiosyncracies within groups or ages - for example, high performing directors might generally wait until they are experienced enough to direct their first film.

Instead we'll plot, for each individual and film, how well they have perfomed relative to their career average (the average tomatometer of all that individual's films). To illustrate this approach let's look at Denzel Washington's films. Denzel has a healthy mean tomatometer of 68%.

<br>
```{r echo=FALSE, results='asis'}

master %>% 
  filter(
    !(role %in% c("actor_4","actor_5", "actor_6")),
    name == "Denzel Washington",
    between(age_at_production, 15,85)
  ) %>%
  group_by(name) %>% 
  mutate(
    average_tomato  = round(mean(tomatometer))
  ) %>% 
  ungroup() %>%
  mutate(
    tomato_index = (tomatometer - average_tomato) / average_tomato
  ) %>%
  sample_n(10) %>%
  arrange(age_at_production) %>%
  select(title, year, age_at_production, tomatometer, average_tomato, tomato_index) %>%
  mutate(tomato_index = scales::percent(tomato_index)) %>%
  datatable(
    style = 'bootstrap',
    class = 'table-hover',
    rownames=FALSE,
    autoHideNavigation = TRUE,
    extensions = 'FixedColumns',
    options = list(
      scrollX = TRUE,
      fixedColumns = list(leftColumns = 1, rightColumns=1)
    )
  )

```
<br>

So based on this 10 film sample Denzel had a solid early 40s where his films (**Hurricane** and **Courage Under Fire** etc) averaged a rotten tomatoes rating about 20% above his career average. He looks like he dropped off after 50 with a few flops like **Déjà Vu** (which I didn't mind) and **2 Guns**. Look I'm not calling him washed but he hasn't done anything good for a while.

@@

To help visualise this approach let's take 12 of the top 20 actors by volume and visualise average their film rating performance by age. De Niro is a great example of someone who starred in increasingly poor films as he got older - clearly visible in the chart below. This probably has more to do with role selection than deteriorating skills (we'll discuss this later) but it's interesting to visualise nonetheless.

@@

## The Upshot

@@

While it appears that actors and directors have their prime between the ages of 27-35 I don't think it's fair to say this is **caused** by a natural skill trajectory. Some caveats that could potentially explain this trend without our skills and abilities narrative:

* Most films typically have non-elderly main characters; limiting the ability for an older, but still skillful, actor to be cast in these roles
* Seperating an Actor's role **choice** vs their **performance** given they chose a particular role highlights the problem we see with De Niro above. It's quite possible that after career success and achievements actor's and directors prioritise different things (eg. *$Cash Money$*) in the twiglights of their careers.

All we can say for sure is that the prime of your acting life (the quality of the films you're in) happens between 27-35. Whether that's becuase you were at the peak of your abilities, or you were more likely to be seeking critically acclaimed roles (hunting that first Oscar), or you were just given opportunities you wouldn't be afforded later in life we'll have to leave to another day.
